---
title: "R Bootcamp: Data Exploration with dplyr"
author: "Paul Jung, Project Mosaic"
date: "June 15, 2017"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, results="hide")
```

## 1. R / R Studio

### Why R?
- R is a computer language specialized in statistical computation
- R Studio is an integrated development environment(IDE) for R
- You can easily understand using R by utilizing R Studio
- You can downlad R / RStudio free through Web!

### Strnegth of R
- Free license(!!!)
- Tremendous pool of users from diverse research fields
- Developing very fast (CRAN)
- Extensive use and flexibility
- Widely used in: Humanities, Social Sciences, Natural Sciences, Business & Industry (Business Analytics)

### R Markdown

- R Script + Research Note
- Increase productivitiy in your data management and analysis
- Write down the step of your analysis in R Markdown
- Code chunck would allow you to track the analysis step-by-step
- link to [R Markdown](https://rmarkdown.rstudio.com/)
- link to R Markdown [Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

## 2. Data Exploration with R - Basics
- [World Bank data](http://data.worldbank.org/) is used for R Bootcamp
- For data manipulation, we Mainly use "dplyr" package
- "dplyr" [Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) offers good guides

### 1) Installing and Loading Package

- Package is a set of functions developed for specific purposes
- e.g., "igraph" is used for network analysis, "haven" is used for importing other forms of datasets, "ggplot2" is used for data visualization
- You can install packages from CRAN repository by web
```{r}
# install.packages("dplyr")
# install.packages("haven")
# install.packages("tidyverse")
```

- You can load packages by "library"-ing it
```{r}
library("dplyr")
library("haven")
```

- Note: ["tidyverse"](http://tidyverse.org/) is a "combination" package for popular packages
```{r}
library("tidyverse")
```

### 2) Setting Working Directory
- Setting working directory makes you reach to files easily
- Working directory is directory where you work with your files
- It makes a shortcut to them
```{r}
getwd() ## print the current working directory

setwd("C:/Users/pjung1/Dropbox (UNC Charlotte)/R Bootcamp") ## set up working directory
# setwd("/Users/pjung1/Dropbox/R Bootcamp") ## For Mac
# setwd("./Users/pjung1/Dropbox/R Bootcamp") ## For Linux

getwd() ## print the (new) current working directory

## Note) For Mac and Linux

```

- In R Markdown, the working directory is set up by the following code
```{r}
# Setting working directory in R Markdown
knitr::opts_knit$set(root.dir = "C:/Users/pjung1/Dropbox (UNC Charlotte)/R Bootcamp")
```

### 3) Importing data
- Import World Bank data (customized for today)
- Population, GDP, health expenditure, land area, mortality rate...
```{r}
# Import csv file
d.POP <- read.csv("C:/Users/pjung1/Dropbox (UNC Charlotte)/R Bootcamp/Data/WB_pop_total.csv")

setwd("C:/RBootcamp/r-bootcamp-workshop-summer-2017-master")

# read_dta: "haven" function to load STATA dta file

library("haven")
d.POP.1 <- read_dta("C:/Users/pjung1/Dropbox (UNC Charlotte)/R Bootcamp/Data/WB_pop_total.dta")
d.POP.2 <- read_dta("Data/WB_pop_total.dta") ## Using set working directory (shortcut)
```

### Practice 1: Import other csv data files from the "data" folder
```{r}
d.GDP <- read.csv("Data/WB_gdp.csv", header = T, stringsAsFactors = F)
d.HEX <- read.csv("Data/WB_healthexp.csv", header = T, stringsAsFactors = F)
d.LAN <- read.csv("Data/WB_landarea.csv", header = T, stringsAsFactors = F)
d.MOR <- read.csv("Data/WB_mortality.csv", header = T, stringsAsFactors = F)

## "header = T" means it will read the first row as column name
## "stringAsFactors = F" means it will consider columns with characters as "characters", not categories

## POP: Population, total
## GDP: GDP (Current $)
## HEX: % of health expenditure to total GDP (%)
## LAN: Land area (sq. km)
## MOR: Mortality rate (per 1000 lives birth)
```

### 4) Reshaping data
- If you look d.POP, you can see each variable means population of each year
- We would like to build "year" column, and stack the records by year
- This job is called "reshaping" wide data to long data
``` {r}
library("tidyr")

# Wide to long: "gather"
gather(d.POP, "year", "POP", 4:18)

d.POP %>% gather("year", "POP", 4:18) ## With "Pipelining", "%>%"

d.POP.L <- d.POP %>%
    gather("year", "POP", 4:18) ## Restoring reshaped data as "d.POP.L"

# Long to wide: "spread"
spread(d.POP.L, year, POP)

d.POP.L %>%
  spread(year, POP)
```


### Practice 2: Reshape the raw data from wide to long
```{r}
d.GDP.L <- d.GDP %>% gather("year", "GDP", 4:18)
d.HEX.L <- d.HEX %>% gather("year", "HEX", 4:18)
d.LAN.L <- d.LAN %>% gather("year", "LAN", 4:18)
d.MOR.L <- d.MOR %>% gather("year", "MOR", 4:18)

d.GDP.L
d.HEX.L
d.LAN.L
d.MOR.L
```

### 5) Dropping variables
- In each data frame, we have unused column, "indicator"
- We have to drop "indicator" column
```{r}
## With Base R Approach
dd <- d.POP.L
dd[, -which(colnames(d.POP.L)=="indicator")]
dd[, -3]

## With "dplyr"
select(d.POP.L, -indicator)
d.POP.L %>%
  select(-indicator)
```

### Practice 3: Dropping variables
```{r}
## Storing data sets without "indicator" column
d.GDP.L <- d.GDP.L %>% select(-indicator)
d.POP.L <- d.POP.L %>% select(-indicator)
d.HEX.L <- d.HEX.L %>% select(-indicator)
d.LAN.L <- d.LAN.L %>% select(-indicator)
d.MOR.L <- d.MOR.L %>% select(-indicator)

d.GDP.L
d.POP.L
d.HEX.L
d.LAN.L
d.MOR.L
```

### 6) Merging data
- We have POP, HEX, LAN, MOR data, but want to merge them as a single data frame
- They should be merged based on country name, region and year
```{r}
## Merge d.POP.L and d.HEX.L first
d <- full_join(d.POP.L, d.GDP.L, by=c("name", "region", "year"))
d

## Merge d and others by making chain
d <- full_join(d, d.HEX.L, by=c("name", "region", "year"))
d <- full_join(d, d.LAN.L, by=c("name", "region", "year"))
d <- full_join(d, d.MOR.L, by=c("name", "region", "year"))
d
```

- Note: Merge function - left_join, right_join, inner_join, full_join

### 7) Setting Variable Type
- variable type in R: atomic, numeric, character, factor, integer, Date, ...
- If you are going to use cateogorical variables (e.g., country, region), it would be better to change them to factor variable
- This makes R to consider them cateogorical variables
```{r}
## To see structure of data frame or see each variable's data type

## With Base R Approach
str(d)

## With "dplyr" package
glimpse(d)
```

```{r}
## Converting to character vairable
d$name <- as.character(d$name)

## Converting to factor variable
d$name <- as.factor(d$name)
d$region <- as.factor(d$region)

## Converting to numeric variable
d$GDP <- as.numeric(d$GDP)
```

### Practice 4: Setting Variable Type
- We want to set up "year" as a numeric variable
```{r}
## Erase "y"
d$year <- gsub("y", "", d$year)
d

## Convert the type to "numeric"
d$year <- as.numeric(d$year)
d
```

### 8) Subsetting by row - filter: Subsetting rows matching criteria
```{r}
## With Base R approach
d[which(d$year==2005 & d$region=="Europe & Central Asia"),]

## With "dplyr" package

## AND: "&" or "," to represent AND condition
filter(d, year==2005 & region=="Europe & Central Asia")
## Make table of observations in 1970 AND in region 6 from "d"

## OR: "|" to represent | condition
filter(d, year==2005 | year==2010 & region=="Europe & Central Asia")
## Make table of observations in 1970 OR 1980 and in region 1 from "d"
```

### Practice 5: Subsetting by row
```{r}
## Defining as an object
d1 <- filter(d, year==2005 & region=="North America")
d2 <- filter(d, year==2005 & region=="North America" | region=="Latin America & Caribbean")
d1
d2
```

### 9) Subsetting by column - select: Pick columns by name
```{r}
## With Base R approach
d[, c("name", "year", "GDP", "HEX")]

## With "dplyr" package
select(d, name, year, GDP, HEX)
## Make table whose columns are name, year, GDP, HEX from d
```

### 10) Subsetting by column and row - %>%: "Chaining" and "Pipelining"
```{r}
## With Base R approach
d[which(d$region=="North America"&d$year<2010), c("name", "region", "year", "POP", "GDP")]

## With "dplyr" package
d %>%
  select(name, region, year, POP, GDP) %>%
  filter(year<2010)
## Among the observations before 2010,
## make table of name, year and POP and GDP
```

### Practice 6: Subsetting by column and row
```{r}
## From observations whose POP is above 1M
## and after 2005 in "Europe & Central Asia"
## make table of GDP and POP

d %>%
  select(name, year, GDP, POP) %>%
  filter(POP>10000000 & year>2005)
```

### 11) Reordering rows - arrange
```{r}
## With Base R Approach
d[order(d$GDP), c("name", "year", "GDP")]

## With "dplyr"
d %>%
  select(name, year, GDP) %>%
  arrange(GDP)
## Among the table of name, year and GDP,
## arrange the tables from the lowest to the highest value of GDP
## with ascending order
```

### Practice 7: Reordering rows
```{r}
## Reordering rows by country name, year and GDP
d %>%
  select(name, year, GDP) %>%
  arrange(name, year, GDP)

```

```{r}
## Note: use 'desc' for descending order
d %>%
  select(name, year, GDP) %>%
  arrange(desc(GDP))
## Among the table of name, year and GDP,
## arrange the tables from the lowest to the highest value of "GDP"
## with descending order
```

### 12) Adding new variable - mutate
```{r}
## With Base R Approach
d$GDP_cap <- d$GDP/d$POP ## GDP_cap: GDP per capita
d$HEX_total <- d$HEX*d$GDP ## HEX_total: Total health expenditure

## With "dplyr"
## To print
d %>%
  select(name:year, GDP, POP, HEX) %>%
  mutate(GDP_cap=GDP/POP, HEX_total=HEX*GDP)

## Log variables
d %>%
  select(name:year, GDP, POP, HEX) %>%
  mutate(lnGDP=log(GDP))

## Other example to store
d4 <- d %>%
  select(name:year, GDP, POP, HEX, LAN) %>%
  mutate(GDP_cap=GDP/POP,
         HEX_total=HEX*GDP,
         lnGDP=log(GDP),
         DEN=POP/LAN)
d4
```

### 13) Renaming Variables
```{r}
## With Base R Approach
names(d)[which(names(d)=="POP")] <- "population"

names(d)[which(names(d)=="population")] <- "POP"

## With "dplyr"
d %>%
  rename(population=POP, area=LAN, health=HEX)
## rename POP -> "population", LAN -> "area", HEX -> "health"
```

### 14) Summarising data table
```{r}
## With Base R Approach
aggregate(POP ~ region + year, d, sum)
## Make table of sum of population by year

## With "dplyr"
d %>%
  group_by(region, year) %>% ## observations will be grouped by year
  summarise(POP_sum=sum(POP, na.rm=T))
## Make table of sum of population by year

t <- d %>%
  group_by(region, year) %>% ## observations will be grouped by region and year
  summarise(POP=sum(POP, na.rm=T),
            GDP=sum(GDP, na.rm=T),
            HEX_total=sum(HEX*GDP, na.rm=T),
            GDP_cap=sum(GDP, na.rm=T)/sum(POP, na.rm=T),
            DEN=sum(POP, na.rm=T)/sum(LAN, na.rm=T))
t
```

### 15) Saving Data as .RData
```{r}
## Saving the whole objects in the environment
save.image("data.RData")

## Saving selected objects
save(d, t, file="table.RData")

## Exporting as CSV file
write.csv(d, "data.csv", row.names=F)
```

## 3. Some Useful functions for Data Exploration

### Random sampling
```{r}
sample_n(d, size=100, replace=T)
sample_n(d %>%
           filter(year==2010), size=40, replace=F)
```

### Frequency Table
```{r}
t2 <- d %>%
  filter(year==2012) %>%
  count(region) %>%
  mutate(prop=prop.table(n))
t2

t3 <- d %>%
  count(region, year) %>%
  mutate(prop=prop.table(n))
t3 
```

### Formatting table in APA format
- Use "apaStyle" package
- Use ["stargazer"](http://jakeruss.com/cheatsheets/stargazer.html) packagefor formatting table neatly
```{r}
# install.packages("apaStyle")
# library("apaStyle")

# APA table
# apa.table(data=t2,
#           level1.header=c("Region", "Number", "Percentage"),
#           title="Frequency table",
#           note="Year is 2012.",
#           file="APA_table.docx")
```
- Note) Other functions for descriptive statistics and regression table
- apa.descriptives, apa.regression (See ["apaStyle"](https://cran.r-project.org/web/packages/apaStyle/apaStyle.pdf))
