---
title: "R Bootcamp: Interactive Visualizations"
author: "Ryan Wesslen"
date: "June 16, 2017"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## Load dataset

Let's read in our dataset.

```{r data}
library(tidyverse)

d <- read_csv("./Data/data.csv")
```

## Review Practice Problems

1.  Create a box plot with the x-axis as year.

First, let's try the obvious plot, using `x = year`, `y = MOR`, and adding `geom_boxplot()`.

However, running this we get a weird result: why?

```{r}
ggplot(d, aes(x = year, y = MOR)) +
  geom_boxplot()
```

This is a formatting issue because our dataset has year as an integer, not a category.

What's an easy way to remediate this problem? Transforming year to a factor.

```{r}
ggplot(d, aes(x = as.factor(year), y = MOR)) +
  geom_boxplot()
```

2.  Next, use a `facet` layer to trellis the box plots by region so that each region has its own temporal box plots. Then create a new plot with only Sub-Saharan Africa and South Asia.

For the next part, we can `facet` to create a plot for each region.

```{r}
ggplot(d, aes(x = as.factor(year), y = MOR, color = region)) +
  geom_boxplot() + facet_wrap(~region)
```

The chart is still too busy. Perhaps we want to look at only Sub-Saharan Africa and South Asia.

```{r}
reg <- c("Sub-Saharan Africa", "South Asia")

ggplot(d[d$region %in% reg,], aes(x = as.factor(year), y = MOR, color = region)) +
  geom_boxplot()
```

3.  Clean up the plot with Sub-Saharan Africa & South Asia by adding x, y, and title labels; moving the legend to the bottom; and color each region by a unique color.

```{r}
ggplot(d[d$region %in% reg,], aes(x = as.factor(year), y = MOR, fill = region)) +
  geom_boxplot() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  labs(x = "Year", y = "Mortality Rate Distribution", title = "Sub-Saharan Africa & South Asia: Mortality Rates") +
  scale_color_discrete(guide = guide_legend(title = "Region"))
```

## themes and ggthemes

Themes can also be helpful to modify the overall style of a plot.

There are helpful add-ins within the `ggthemes` package, which can provide a broader range of templates.

For example: [Tufte](https://www.edwardtufte.com/tufte/) Box Plots ("Minimum Ink")

First, let's rerun the data manipulations we did yesterday...

```{r}
pop2014 <- d[d$year==2014,]
pop2014$region <- factor(pop2014$region, 
                         levels = c("Sub-Saharan Africa",
                                    "South Asia",
                                    "East Asia & Pacific",
                                    "Latin America & Caribbean",
                                    "Middle East & North Africa",
                                    "Europe & Central Asia",
                                    "North America"
                                    ))

levels(pop2014$region) <- gsub("&", "\n &", levels(pop2014$region))
```

Now, let's create the plot (after calling `ggthemes`)...

```{r}
#install.packages("ggthemes")
library(ggthemes)

g <- ggplot(pop2014, aes(region, MOR)) + labs(x = "Region", y = "Mortality Rate") 
g 
```

change the theme...

```{r}
g + theme_tufte(ticks=FALSE)
```

then add the geom...

```{r}
g + theme_tufte(ticks=FALSE) +
  geom_tufteboxplot()
```

or modify it...

```{r}
g + theme_tufte(ticks=FALSE) +
  geom_tufteboxplot(median.type = "line", whisker.type = 'line')
```

```{r}
g + theme_tufte(ticks=TRUE) +
  geom_tufteboxplot(median.type = "line", whisker.type = 'line', hoffset = 0, width = 2)
```

There are many other themes like...

```{r}
#recall from yesterday's code
years <- c(2001, 2005, 2009, 2013)
d1 <- d[d$year %in% years,]

p <- ggplot(d1, aes(x = HEX, y = MOR, color = region)) +
  geom_point() +
  facet_wrap(~year) + 
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Health Expenditure as a Percent of GDP (Log)", y = "Mortality Rate (Log)") +
  geom_smooth(method=lm, se=FALSE)  # Don't add shaded confidence region

p + theme_solarized() +
  scale_colour_solarized("blue") +
  scale_color_discrete(guide = guide_legend(title = "Region"))
```

```{r}
p + theme_igray()
```

```{r}
p + theme_few() + 
  scale_colour_few() + 
  scale_color_discrete(guide = guide_legend(title = "Region"))
```

Here's a full list: [https://github.com/jrnold/ggthemes](https://github.com/jrnold/ggthemes)

## ggplot2 for plotly

Plotly is a package of interactive visualizations.

In fact, they have a `ggplot2` package called [`plotly`](https://plot.ly/ggplot2/) that is a wrapper function `ggplotly` on `ggplot2` but for interactive visualizations.

There are multiple ways to run interactive `ggplot`, but the easiest is to use the same logic for `ggplot` but then pass the object to the function `ggplotly()`.


```{r}
#install.packages("plotly")
library(plotly)

ggplotly(p)
```

```{r}
#rerun data manipulations from yesterday
d$r.MOR <- d$MOR * (d$POP) / 1000
d2 <- d %>% 
  na.omit() %>% # note the na.omit is very important!
  group_by(region, year) %>% 
  summarise(t.MOR=sum(r.MOR),
            t.POP=sum(POP)) %>%
  mutate(MOR.rate = t.MOR / (t.POP / 1000))

p <- ggplot(d2, aes(year, MOR.rate, color = region)) + 
  geom_line() +
  geom_point() + 
  labs(x = "Year", y = "Mortality Rate") +
  scale_color_discrete(guide = guide_legend(title = "Region"))

ggplotly(p)
```

## Web-interactive Visualizations

What's the range of possible ggplot2 options?

There's a great open source Shiny app to explore ggplot2: [https://www.r-bloggers.com/shiny-app-to-explore-ggplot2/](https://www.r-bloggers.com/shiny-app-to-explore-ggplot2/)

This visualization utilizes [Shiny](https://shiny.rstudio.com/), which is a R visualization toolkit to build interactive visualizations without knowing JavaScript, HTML, and CSS.

I've previously created a [R Visualization Tutorial](https://github.com/wesslen/RVisTutorial) that includes several examples using Twitter data.